Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-22 concl.tex
Liz is done rewriting

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of a time
series.  To operationalize that definition, we use an approximation of
the Kolmogorov-Sinai entropy~\cite{lind95}, estimated using a weighted
version of the permutation entropy of~\cite{bandt2002per}.  This WPE
technique---an ordinal calculation of forward information transfer in
a time series---is ideal for our purposes because it works with
real-valued data and is known to converge to the true entropy value.
Using a variety of forecast models and more than a hundred time-series
data sets from a computer-performance experiment, we have shown that
prediction accuracy is indeed correlated with weighted permutation
entropy: the higher the WPE, in general, the higher the prediction
error.  There are some exceptions: \svdone is easier to predict than
its WPE value suggests using any of the three forecast strategies
studied here.  Conversely, forecast errors for the \col and \svdfive
signals are higher than the corresponding WPEs suggest---except for
the nonlinear LMA predictor.

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  (This is particularly apparent
in the \col and \svdfive examples.)
% , where the other two methods do not perform well.)  
The \naive ~method, which simply predicts the mean, can work very well
on noisy signals because it effects a filtering operation.  The simple
random-walk strategy outperforms LMA, ARIMA, and the \naive ~method on
the \gcc signal, which is extremely complex---i.e., extremely low
redundancy.
%  \naive ~wins on \svdone with the exception to the 5 outliers.
The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} operationalizes the discussion in this
paragraph.  It is a useful empirical guideline for knowing when a
model is not well-matched to the task at hand: a MASE score that is
more than $WPE = 0.2794 \cdot MASE + 0.5334$
times the WPE of the corresponding signal is an indication that that
time series has more predictive structure than that forecast model can
capture and exploit.

Given that information is a fundamental limitation in predictability,
then gathering and using more information is an obvious next step.
But there is an equally obvious tension here between data length and
prediction speed: a forecast that requires half a second to compute is
not useful for the purposes of real-time control of a computer system
with a MHz clock rate.  Another alternative is to sample several
system variables simultaneously and build multivariate models.  This
is a particular challenge in nonlinear LMA-type models, since
multivariate delay-coordinate embedding (e.g.,
\cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
computationally prohibitive.  We are working on alternative methods
that sidestep that complexity.

@@

All of the experiments described in this paper involve a particular
physical system.  We chose that system because its behavior spans a
broad spectrum of classical time series patterns.  Since these
patterns---dynamical chaos, periodicities, noise, etc.--- are generic,
we believe that our results will generalize beyond this particular
system.  Testing that claim with different examples is another
important next step.  Of particular interest would be the class of
so-called \emph{hybrid systems}: [[insert def]].  A lathe, for
instance, that has an intermittent instability.  Traffic at an
internet router, which has characteristic patterns during normal
operation that shift radically during an attack.  Modeling and
predicting the behavior of these kinds of systems effectively is quite
difficult; doing so adaptively and automatically---in the manner that
is alluded to at the end of the previous paragraph---would be an even
more important challenge.

>>>>

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of a time
series.  To operationalize that definition, we use an approximation of
the Kolmogorov-Sinai entropy~\cite{lind95}, estimated using a weighted
version of the permutation entropy of~\cite{bandt2002per}.  This WPE
technique---an ordinal calculation of forward information transfer in
a time series---is ideal for our purposes because it works with
real-valued data and is known to converge to the true entropy value.
Using a variety of forecast models and more than a hundred time-series
data sets from a computer-performance experiment, we have shown that
prediction accuracy is indeed correlated with weighted permutation
entropy: the higher the WPE, in general, the higher the prediction
error.  There are some exceptions: \svdone is easier to predict than
its WPE value suggests using any of the three forecast strategies
studied here.  Conversely, forecast errors for the \col and \svdfive
signals are higher than the corresponding WPEs suggest---except for
the nonlinear LMA predictor.  On the whole, though, signals with
higher weighted permutation entropy are harder to predict.

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  (This is particularly apparent
in the \col and \svdfive examples.)
% , where the other two methods do not perform well.)  
The \naive ~method, which simply predicts the mean, can work very well
on noisy signals because it effects a filtering operation.  The simple
random-walk strategy outperforms LMA, ARIMA, and the \naive ~method on
the \gcc signal, which is extremely complex---i.e., extremely low
redundancy.
%  \naive ~wins on \svdone with the exception to the 5 outliers.
The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} operationalizes the discussion in this
paragraph.  It is a preliminary, but potentially useful, empirical
guideline for knowing when a model is not well-matched to the task at
hand: a MASE score that is more than $3.58 \cdot \mathrm{WPE} - 1.91$,
where $\mathrm{WPE}$ is that of the corresponding signal, is an
indication that that time series has more predictive structure than
that forecast model can capture and exploit.

Given that information is a fundamental limitation in predictability,
then gathering and using more information is an obvious next step.
But there is an equally obvious tension here between data length and
prediction speed: a forecast that requires half a second to compute is
not useful for the purposes of real-time control of a computer system
with a MHz clock rate, for instance.  Another alternative is to sample
several system variables simultaneously and build multivariate models.
This is a particular challenge in nonlinear LMA-type models, since
multivariate delay-coordinate embedding (e.g.,
\cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
computationally prohibitive.  We are working on alternative methods
that sidestep that complexity.

@@

All of the experiments described in this paper involve a particular
physical system.  We chose that system because its behavior spans a
broad spectrum of classical time series patterns.  Since these
patterns (dynamical chaos, periodicities, noise, etc.) are generic, we
believe that our results will generalize beyond this particular
system.  Testing that claim with different examples is another
important next step---and one that will be critical to validating our
claims about the utility of the dashed line as a heuristic for
determining whether a particular forecast method is well-matched to a
particular task.  Of particular interest would be the class of
so-called \emph{hybrid systems}, which exhibit discrete transitions
between different continuous regimes---e.g., a lathe that has an
intermittent instability or traffic at an internet router, which has
characteristic patterns during normal operation that shift radically
during an attack.  Effective modeling and prediction of these kinds of
systems effectively is quite difficult; doing so adaptively and
automatically---in the manner that is alluded to at the end of the
previous paragraph---would be an even more important challenge.

### 2014-04-22 intro.tex
added sentence to abstract
Liz is done rewriting

Complicated time-series data are ubiquitous in modern scientific
research.  The complexity of these data spans a wide range.  On the
low end of this spectrum are time series that exhibit perfect
predictive structure, i.e, signals whose future values can be
successfully predicted from past values.  This can be viewed as the
output of an underlying process that generates information and/or
transmits it from the past to the future in a perfectly predictable
fashion.  Constant or periodic signals, for example, fall in this
class.  On the opposite end of this spectrum are signals that are what
we call \emph{fully complex}, where the underlying generating process
transmits no information at all from the past to the future.  White
noise and random-walk processes fall in this class.  In fully complex
signals, knowledge of the past gives no insight into the future,
\emph{regardless} of what model one chooses to use.  Signals in the
midrange of this spectrum---e.g., deterministic chaos---pose
interesting challenges from a modeling perspective.  In these signals,
enough information is being transmitted from the past to the future
that an \emph{ideal} model---one that captures the generating
process---can forecast the future behavior of the observed system with
high accuracy.

@@

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation~\cite{crutchfield2003}.  
\label{page:redundancy}
Our approach exploits this decomposition in order to assess how much
predictive structure is present in a signal---i.e., where it falls on
the complexity spectrum mentioned above.  We define \emph{complexity}
as a particular approximation of Kolmogorov-Sinai
entropy~\cite{lind95}.  That is, we view a random-walk time series
(which exhibits high entropy) as purely complex, whereas a periodic
signal, which exhibits low entropy, is on the low end of the
complexity spectrum.  We argue that an extension of \emph{permutation
  entropy}~\cite{bandt2002per}---a method for approximating the
entropy through ordinal analysis---is an effective way to assess the
complexity of a given time series.  Permutation entropy is ideal for
our purposes because it works with real-valued data and is known to
converge to the true entropy value. Other existing techniques either
require specific knowledge of the generating process or produce biased
values of the entropy~\cite{bollt2001}.

We focus on real-valued, scalar, time-series data that appear\alert{s}
complicated: i.e., not periodic, linear, constant, etc.  We make no
assumptions about the properties of the generating process: whether it
is linear, nonlinear, deterministic, stochastic, stationary,
non-stationary, etc.
To explore the relationship between complexity, predictive structure,
and predictability, we generate forecasts for a variety of time-series
datasets using four different prediction methods, then compare the
accuracy of those predictions to the permutation entropy of the
associated signals.  This results in two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is {\sl (i)}
  quantifiable by permutation entropy and {\sl (ii)} correlated with prediction
  accuracy of an appropriate predictor.
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema---and in the choice of which one is appropriate for a given 
time series.
\end{enumerate}
The forecast methods used in this study were chosen to be a
representative sampling of standard prediction strategies, but they do
not, of course, cover that space exhaustively.  Our goal here is an
empirical assessment of the relationship between predictability and
complexity, not formal results about a ``best'' predictor for a given
time series.  From a practitioner's standpoint, it would be useful to
know {\sl a priori} if a time series contains enough predictive structure to
make it worth spending the time and effort searching for a good
forecast method.  It would also be useful to know if a given method is
inadequate---that is, if an alternative method could do better and so 
one should continue searching.

@@

For the purposes of this study, we require a broad array of
time-series datasets from across the complexity spectrum.  Rather than
use synthetic data, we study sensor data from computer performance
experiments.  While this is not a common laboratory experiment, it is
a highly appropriate choice.  Computers are extremely complicated
systems.  Modern microprocessor chips contain multiple processing
units and multi-layer memories, for instance, and they use complicated
hardware/software strategies to maximize performance by moving data
and threads of computation across those resources.  As a result, the
processor and memory loads during the execution of even a very simple
program can evolve in a very complicated fashion.
Figure~\ref{fig:col-ipc} shows a performance trace of a three-line
loop that repeatedly initializes the upper triangle of a matrix in
column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic~\cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the code can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making them an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting computer \emph{performance}, which does not violate
  the Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

>>>>

Complicated time-series data are ubiquitous in modern scientific
research.  The complexity of these data spans a wide range.  On the
low end of this spectrum are time series that exhibit perfect
predictive structure, i.e, signals whose future values can be
successfully predicted from past values.  Signals like this can be
viewed as the product of an underlying process that generates
information and/or transmits it from the past to the future in a
perfectly predictable fashion.  Constant or periodic signals, for
example, fall in this class.  On the opposite end of this spectrum are
signals that are what we call \emph{fully complex}, where the
underlying generating process transmits no information at all from the
past to the future.  White noise and random-walk processes fall in
this class.  In fully complex signals, knowledge of the past gives no
insight into the future, regardless of what model one chooses to use.
Signals in the midrange of this spectrum---e.g., deterministic
chaos---pose interesting challenges from a modeling perspective.  In
these signals, enough information is being transmitted from the past
to the future that an \emph{ideal} model---one that captures the
generating process---can forecast the future behavior of the observed
system with high accuracy.

@@

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation~\cite{crutchfield2003}.  
\label{page:redundancy}
Our approach exploits this decomposition in order to assess how much
predictive structure is present in a signal---i.e., where it falls on
the complexity spectrum mentioned above.  We define \emph{complexity}
as a particular approximation of Kolmogorov-Sinai
entropy~\cite{lind95}.  That is, we view a random-walk time series
(which exhibits high entropy) as purely complex, whereas a low-entropy
periodic signal is on the low end of the complexity spectrum.  We
argue that an extension of \emph{permutation
  entropy}~\cite{bandt2002per}---a method for approximating the
entropy through ordinal analysis---is an effective way to assess the
complexity of a given time series.  Permutation entropy is ideal for
our purposes because it works with real-valued data and is known to
converge to the true entropy value. Other existing techniques either
require specific knowledge of the generating process or produce biased
values of the entropy~\cite{bollt2001}.

We focus on real-valued, scalar, time-series data.
% 
% that appears complicated: i.e., not periodic, linear, constant, etc.
% 
We make no assumptions about the properties of the generating process:
whether it is linear, nonlinear, deterministic, stochastic,
stationary, non-stationary, etc.  To explore the relationship between
complexity, predictive structure, and predictability, we generate
forecasts for a variety of time-series datasets using four different
prediction methods, then compare the accuracy of those predictions to
the permutation entropy of the associated signals.  This results in
two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is \emph{(i)}
  quantifiable by permutation entropy and \emph{(ii)} correlated with
  prediction accuracy of an appropriate predictor.
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema---and in the choice of which one is appropriate for a given 
time series.
\end{enumerate}
The forecast methods used in this study were chosen to be a
representative sampling of standard prediction strategies, but they do
not, of course, cover that space exhaustively.  Our goal here is an
empirical assessment of the relationship between predictability and
complexity, not formal results about a ``best'' predictor for a given
time series.  From a practitioner's standpoint, it would be useful to
know \emph{ a priori} if a time series contains enough predictive
structure to make it worth spending the time and effort searching for
a good forecast method.  It would also be useful to know if a given
method is inadequate---that is, if an alternative method could do
better and so one should continue searching.

@@

For the purposes of this study, we require a broad array of
time-series datasets from across the complexity spectrum.  Since
forecasting is a real-world problem, we chose to study sensor data
from a computer-performance experiment.  While this is not a common
laboratory experiment, it is a highly appropriate choice here.
Computers are extremely complicated systems.  Modern microprocessor
chips contain multiple processing units and multi-layer memories, for
instance, and they use complicated hardware/software strategies to
maximize performance by moving data and threads of computation across
those resources.  As a result, the processor and memory loads during
the execution of even a very simple program can evolve in a very
complicated fashion.  Figure~\ref{fig:col-ipc} shows a performance
trace of a short program that repeatedly initializes the upper
triangle of a matrix in column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a simple C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic~\cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the code can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making them an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting computer \emph{performance}, which does not violate
  the Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

### 2014-04-22 methods.tex
Liz is done rewriting

The time-series data for these experiments was collected on an Intel
Core\textsuperscript{\textregistered} i7-2600-based machine running
the 2.6.38-8 Linux kernel.  This particular microprocessor chip has
eight processing units, a clock rate of 3.40 GHz, and a cache size of
8192 KB.  We gathered performance traces during the execution of three
different programs---the simple \col loop whose performance is
depicted in Figure~\ref{fig:col-ipc}, as well as two more-complex
programs: one from the SPEC 2006CPU benchmark suite (\gcc), and one
from the LAPACK linear algebra package (\svd).  In these experiments, the scalar observation $x_i$ was a measurement of the processor performance at time $i$ during the execution of each program.  To record these measurements, we used the {\tt libpfm4}
library, via PAPI
% 
% \footnote{Performance Application Programming Interface}
% 
5.2~\cite{papi}, to stop program execution at 100,000-instruction
intervals and read the contents of the CPU's onboard hardware
performance monitors\footnote{These specialty registers are built into
  modern microprocessors for the purpose of on-board monitoring and
  storage of performance data.  They can be programmed to measure a
  wide array of processor and memory metrics besides the one used
  here.}, which we had programmed to count how many instructions were
executed in each clock cycle (IPC).  For an in-depth description of
this custom-measurement infrastructure, including discussion of the
implications of the sampling interval, please see
\cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical validation,
we collected 15 performance traces from each of the three programs.
These traces, and the processes that generated them, are described in
more depth in the rest of this section.

@@

The SPEC CPU2006 benchmark suite \cite{spec} is a collection of
complicated programs that are used in the computer science community
to assess and compare the performance of different computers.  \gcc is
a member of that suite.
% 
% This program is written by Richard Stallman and is based on Version
% 3.2 of {\tt gcc}. This benchmark essentially runs as a compiler with
% many of its optimization flags enabled, compiling a series of input
% files and generating x86-64 assembly code files intended for execution
% on AMD Opteron processors\cite{spec}. 
% 
It is a \emph{C-compiler}: a program that translates code written in the
C language into a lower-level format that can be executed by the
processor chip.  Its behavior is far more complicated than that of
\col, as is clear from Figure~\ref{fig:gcc-ts}.
  \begin{figure}[t]
  \centering
    \includegraphics[width=\columnwidth]{figs/gccfullts}
    \caption{The instructions per CPU clock cycle (IPC) during the
      execution of \gcc. Each point is the average IPC in a 100,000
      instruction period.}
    \label{fig:gcc-ts}
  \end{figure}
Unlike \col, where the processor utilization is quite structured,
\gcc's performance appears almost random.

\svd is a Fortran program from the LAPACK linear algebra package
\cite{lapack}.  It calculates the singular value decomposition of a
rectangular $M$ by $N$ matrix with real-valued entries.  For our experiments, we
choose $M=750$ and $N=1000$ and generate the matrix entries randomly.
% 
% \footnote{Multiple randomly generated matrices were
%   investigated but no measurable effect was present in the resulting
%   time series.}. 
% 
The behavior of this program as it computes the singular values of
this matrix is very interesting, as shown in Figure
\ref{fig:svd-ts-colored}.  As the code moves though its different
phases---diagonalizing the matrix, computing its transpose,
multiplying, etc.---the processor utilization changes quite radically,
as shown in Figure~\ref{fig:svd-ts-colored}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
    \caption{The instructions per CPU clock cycle (IPC) during the
      execution of \svd. Each point is the average IPC in a 100,000
      instruction period.  The colors identify the different {\sl
        segments} of the signal that are discussed in the text.}
    \label{fig:svd-ts-colored}
  \end{figure}
For the first $\sim$21,000 measurements (21,000 $\times$ 100,000
instructions), roughly 1.8 instructions are executed per cycle, on the
average, by the eight processing units on this chip.  After that, the
IPC moves through a number of different oscillatory regimes, which we
have color-coded in the figure in order to make textual
cross-references easy to track and understand.

>>>>

The time-series data sets for these experiments were collected on an
Intel Core\textsuperscript{\textregistered} i7-2600-based machine
running the 2.6.38-8 Linux kernel.  This particular microprocessor
chip has eight processing units, a clock rate of 3.40 GHz, and a cache
size of 8192 KB.  We gathered performance traces during the execution
of three different programs---the simple \col loop whose performance
is depicted in Figure~\ref{fig:col-ipc}, as well as two more-complex
programs: one from the SPEC 2006CPU benchmark suite (\gcc), and one
from the LAPACK linear algebra package (\svd).  In these experiments,
the scalar observation $x_i$ was a measurement of the processor
performance at time $i$ during the execution of each program.  To
record these measurements, we used the {\tt libpfm4} library, via PAPI
% 
% \footnote{Performance Application Programming Interface}
% 
5.2~\cite{papi}, to stop program execution at 100,000-instruction
intervals---the unit of time in this paper---and read the contents of
the CPU's onboard hardware performance monitors\footnote{These
  specialty registers are built into modern microprocessors for the
  purpose of monitoring and storage of performance data.  They can be
  programmed to measure a wide array of processor and memory metrics
  besides the one used here.}, which we had programmed to count how
many instructions were executed in each clock cycle (IPC).  For an
in-depth description of this custom-measurement infrastructure,
including discussion of the implications of the sampling interval,
please see \cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical
validation, we collected 15 performance traces from each of the three
programs.  These traces, and the processes that generated them, are
described in more depth in the rest of this section.

@@

The SPEC CPU2006 benchmark suite \cite{spec} is a collection of
complicated programs that are used in the computer-science community
to assess and compare the performance of different computers.  \gcc is
a member of that suite.
% 
% This program is written by Richard Stallman and is based on Version
% 3.2 of {\tt gcc}. This benchmark essentially runs as a compiler with
% many of its optimization flags enabled, compiling a series of input
% files and generating x86-64 assembly code files intended for execution
% on AMD Opteron processors\cite{spec}. 
% 
It is a \emph{compiler}: a program that translates code written in a
high-level language (C, in the case of \gcc) into a lower-level format
that can be executed by the processor chip.  Its behavior is far more
complicated than that of \col, as is clear from
Figure~\ref{fig:gcc-ts}.
  \begin{figure}[t]
  \centering
    \includegraphics[width=\columnwidth]{figs/gccfullts}
    \caption{The instructions per CPU clock cycle (IPC) during the
      execution of \gcc. Each point is the average IPC in a 100,000
      instruction period.}
    \label{fig:gcc-ts}
  \end{figure}
Unlike \col, where the processor utilization is quite structured,
\gcc's performance appears almost random.

\svd is a Fortran program from the LAPACK linear algebra package
\cite{lapack}.  It calculates the singular value decomposition of a
rectangular $M$ by $N$ matrix with real-valued entries.  For our experiments, we
choose $M=750$ and $N=1000$ and generate the matrix entries randomly.
% 
% \footnote{Multiple randomly generated matrices were
%   investigated but no measurable effect was present in the resulting
%   time series.}. 
% 
The behavior of this program as it computes the singular values of
this matrix is very interesting, as shown in Figure
\ref{fig:svd-ts-colored}.  As the code moves though its different
phases---diagonalizing the matrix, computing its transpose,
multiplying, etc.---the processor utilization patterns change quite radically,
as shown in Figure~\ref{fig:svd-ts-colored}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
    \caption{The instructions per CPU clock cycle (IPC) during the
      execution of \svd. Each point is the average IPC in a 100,000
      instruction period.  The colors identify the different \emph{
        segments} of the signal that are discussed in the text.}
    \label{fig:svd-ts-colored}
  \end{figure}
For the first $\sim$21,000 measurements (21,000 $\times$ 100,000
instructions), roughly 1.8 instructions are executed per cycle, on the
average, by the eight processing units on this chip.  After that, the
IPC moves through a number of different oscillatory regimes, which we
have color-coded in the figure in order to make textual
cross-references easy to track.
